Density-dependent selection and the limits of relative fitness

### 2018-03-19 manuscript.tex
latest

In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic'' growth rate (for concreteness we use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic population growth rates. However, growth cannot continue at a constant rate indefinitely. As population density increases, crowding will cause the Malthusian parameters $r_i$ to decline (e.g. \citealt[pp. 203]{begon_1990}). It is then possible that selection is density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}. We would then need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}. In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise a similar problem, because a complete description of selection still only requires us to model the type frequencies, not the ecological variable $N$ as well.

Many previous studies of selection in density-regulated populations have focused on the evolution of traits that affect the ability to keep growing at higher densities  \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}. This focus stems in part from Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, lande_2009}, and MacArthur's concept of $K$-selection \citep{macarthur_1962}. But the principle that density changes in response to selection is also mathematically ingrained in our models of density-regulated population growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), similar ``toy'' models with more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}), and the Lotka-Volterra competition model (at least during a sweep, except in special cases; \citealt{gill_1974,smouse_1976,mallet_2012}). This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps.

In practice, many population genetic models simply ignore density-dependent selection, and and each type is assigned a constant relative fitness. Density either does not enter the model at all, or if finite-population size effects (``drift'') are important, then $N$ is assumed to have reached some fixed equilibrium value. Selection is typically assumed to operate through viability, but the ecological processes underlying the regulation of population density are left unspecified (e.g. \citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}). It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.

@@

However, the preceding arguments do not imply that the constant-$s$ approximation \textit{only} applies under weak selection and stable $N$. Every model of density-regulated growth mentioned above is missing an important aspect of selection which has ramifications for the constant-$s$ approximation: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles. Although similar ideas featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}),

>>>>

In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic'' growth rate (for concreteness we use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic population growth rates. However, growth cannot continue at a constant rate indefinitely. As population density increases, crowding will cause the Malthusian parameters $r_i$ to decline (e.g. \citealt[pp. 203]{begon_1990}). It is then possible that selection is density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}. We would then need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}. In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, in the presence of density-dependent selection, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise a similar problem, because a complete description of selection still only requires us to model the type frequencies, not the ecological variable $N$ as well.

In practice, many population genetic models simply assign a constant relative fitness to each type, and ignore density dependence. Selection is typically interpreted as operating through viability, but the ecological processes underlying the regulation of population density are left unspecified (e.g. \citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}). Density either does not enter the model at all, or if finite-population size effects (``drift'') are important, then $N$ is assumed to have reached some fixed equilibrium value (Fig.~\ref{fig:Ksel}b).

A rather different picture emerges in more ecologically-explicit studies of selection in density-regulated populations. A major focus of previous studies in this area is the evolution of traits that affect the ability to keep growing at higher densities \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}. This focus stems in part from Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, lande_2009}, and MacArthur's concept of $K$-selection \citep{macarthur_1962}. But more broadly, density changes in response to selection in many models of density-regulated growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), similar ``toy'' models with more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}), and the Lotka-Volterra competition model (at least during a sweep, except in special cases; \citealt{gill_1974,smouse_1976,mallet_2012}). This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps.

@@

The preceding arguments do not imply that the constant-$s$ approximation \textit{only} applies under weak selection and stable $N$. Every model of density-regulated growth mentioned above is missing an important aspect of selection which has ramifications for the constant-$s$ approximation: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles. Although similar ideas featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}),
